---
title: "Gather data with BQ"
format: 
    gfm:
      echo: true
      message: false
      warning: false
      code-fold: true
---

## About

This notebook documents steps taken to obtain the publication volume and corresponding market share from commercial publishers using open metadata.
The work is carried out using Google BigQuery.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(bigrquery)
library(DBI)
library(countrycode)
library(rvest)

# Connect to GCBQ from SUB Göttingen
bq_con <- dbConnect(
  bigrquery::bigquery(),
  project = "subugoe-collaborative",
  dataset = "openalex",
  billing = "subugoe-collaborative"
)
```

## Create publication dataset

NB author position is not filled in SUB Göttignen OpenAlex Walden snapshot. 
I infer by author position instead.

```{sql, connection = "bq_con"}
CREATE OR REPLACE TABLE `subugoe-closed.apc_science.md_raw` AS (
SELECT DISTINCT
    oalex.id,
    oalex.doi,
    publication_year,
    publication_date,
    primary_location.source.issn_l AS issn_l,
    oalex.primary_location.source.host_organization_name AS oalex_publisher,
    cr.publisher AS cr_publisher,
    primary_location.source.is_in_doaj AS is_in_doaj,
    open_access.oa_status AS oa_status,
    -- First Author
    country AS country_code,
    inst.ror
  FROM `subugoe-collaborative.openalex_walden.works` AS oalex
  LEFT JOIN
    UNNEST(authorships) AS au WITH OFFSET AS pos
  LEFT JOIN
    UNNEST(au.countries) AS country
  LEFT JOIN
    UNNEST(au.institutions) AS inst  
  LEFT JOIN `subugoe-collaborative.resources.document_classification_september25` as doctype_classifier ON oalex.doi = doctype_classifier.doi
  -- Obtain publisher info from Crossref
  INNER JOIN `subugoe-collaborative.cr_instant.snapshot` as cr ON LOWER(oalex.doi) = LOWER(cr.doi)
  WHERE
      pos = 0  -- first author only
    AND
      primary_location.source.type = "journal"
    AND
      is_paratext = FALSE
    AND 
       oalex.type IN ('article','review')
    AND
       is_research IS NOT FALSE
    AND 
      is_xpac = FALSE
    AND (
      NOT REGEXP_CONTAINS(oalex.biblio.issue, '^[a-zA-Z]')
      OR oalex.biblio.issue IS NULL)
    AND (
      NOT (
        REGEXP_CONTAINS(
          oalex.title,
          '[0-9]{3} pp.')))
    AND publication_year BETWEEN 2018 AND 2025
)
```

Create table with articles enabled by an TA

```{sql}
#| connection: "bq_con"
CREATE OR REPLACE TABLE `subugoe-closed.apc_science.jct_articles` AS (
WITH
  -- Part 1: Enrich ROR variants from associated institutions
  obtain_associated_ror_ids AS (
    SELECT
      esac_id,
      jct_inst.ror_id AS ror_jct,
      inst.ror AS ror_associated
    FROM
      `subugoe-collaborative.openbib.jct_institutions` AS jct_inst
    LEFT JOIN
      `subugoe-collaborative.openalex.institutions` AS oalex_inst
    ON
      jct_inst.ror_id = oalex_inst.ror
    LEFT JOIN
      UNNEST(oalex_inst.associated_institutions) AS inst
  ),

  create_matching_table AS (
    SELECT esac_id, 'ror_jct' AS ror_type, ror_jct AS ror
    FROM obtain_associated_ror_ids
    UNION ALL
    SELECT esac_id, 'ror_associated' AS ror_type, ror_associated AS ror
    FROM obtain_associated_ror_ids
  ),

  enriched_ror_variants AS (
    SELECT DISTINCT
      create_matching_table.*,
      DATE(jct_inst.start_date) AS start_date,
      DATE(jct_inst.end_date) AS end_date
    FROM create_matching_table
    INNER JOIN `subugoe-collaborative.openbib.jct_esac` AS jct_inst
      ON create_matching_table.esac_id = jct_inst.id
  ),

  journal_agreements AS (
    SELECT
      j.issn_l,
      e.id AS esac_id
    FROM `subugoe-collaborative.openbib.jct_journals` AS j
    INNER JOIN `subugoe-collaborative.openbib.jct_esac` AS e
      ON j.esac_id = e.id
  )

SELECT DISTINCT
  md.id,
  md.doi,
  md.issn_l AS matching_issn_l,
  md.ror AS matching_ror,
  erv.ror_type,
  erv.esac_id,
  erv.start_date,
  erv.end_date,
  md.publication_date
FROM `subugoe-closed.apc_science.md_raw` AS md
INNER JOIN journal_agreements AS ja
  ON md.issn_l = ja.issn_l
INNER JOIN enriched_ror_variants AS erv
  ON md.ror = erv.ror
  AND ja.esac_id = erv.esac_id
WHERE
  md.ror IS NOT NULL
  AND md.issn_l IS NOT NULL
  AND (PARSE_DATE('%Y-%m-%d', md.publication_date) >= erv.start_date OR erv.start_date IS NULL)
  AND (PARSE_DATE('%Y-%m-%d', md.publication_date) <= erv.end_date OR erv.end_date IS NULL)
  -- Only TA OA articles
  AND oa_status IN ('gold', 'hybrid')
ORDER BY erv.esac_id, md.publication_date
)
```

Create publisher-level table by publication_year.

```{sql}
#| connection: bq_con
#| output.var: top_20_all
WITH top_20 AS (
  SELECT cr_publisher, COUNT(DISTINCT doi) AS total_articles
  FROM `subugoe-closed.apc_science.md_raw`
  GROUP BY cr_publisher
  ORDER BY total_articles DESC
  LIMIT 20
),
labeled AS (
  SELECT 
    doi,
    publication_year,
    CASE WHEN t.cr_publisher IS NOT NULL THEN m.cr_publisher ELSE 'Other' END AS cr_publisher
  FROM `subugoe-closed.apc_science.md_raw` m
  LEFT JOIN top_20 t USING (cr_publisher)
)
SELECT COUNT(DISTINCT doi) as n_articles, cr_publisher, publication_year
FROM labeled
GROUP BY cr_publisher, publication_year
ORDER BY 
  CASE WHEN cr_publisher = 'Other' THEN 1 ELSE 0 END,  -- Other to bottom
  SUM(COUNT(DISTINCT doi)) OVER (PARTITION BY cr_publisher) DESC,
  publication_year DESC
```

```{r}
top_20_all
write_csv(top_20_all, "data/top_20_all.csv")
```

Create publisher-level table by publication_year, oa_business_model.

```{sql}
#| connection: bq_con
#| output.var: top_20_oa
WITH top_20 AS (
  SELECT cr_publisher, COUNT(DISTINCT doi) AS total_articles
  FROM `subugoe-closed.apc_science.md_raw`
  GROUP BY cr_publisher
  ORDER BY total_articles DESC
  LIMIT 20
),
labeled AS (
  SELECT 
    m.doi,
    m.publication_year,
    m.oa_status,
    CASE WHEN t.cr_publisher IS NOT NULL THEN m.cr_publisher ELSE 'Other' END AS cr_publisher,
    CASE
      WHEN m.is_in_doaj = TRUE OR m.cr_publisher IN ('MDPI AG', 'Frontiers Media SA', 'Public Library of Science (PLoS)') THEN 'Full OA' ELSE 'Hybrid'
    END AS journal_business_model,
    CASE WHEN jct.doi IS NOT NULL THEN TRUE ELSE FALSE END AS enabled_ta
  FROM `subugoe-closed.apc_science.md_raw` m
  LEFT JOIN top_20 t USING (cr_publisher)
  LEFT JOIN (SELECT DISTINCT doi FROM `subugoe-closed.apc_science.jct_articles`) AS jct
    USING (doi)
)
SELECT 
  cr_publisher, 
  publication_year,
  journal_business_model,
  COUNT(DISTINCT doi) AS n_articles, 
  COUNT(DISTINCT CASE WHEN oa_status IN ('gold', 'hybrid') THEN doi END) AS n_oa_articles,
  COUNT(DISTINCT CASE WHEN enabled_ta THEN doi END) AS n_ta_articles
FROM labeled
GROUP BY cr_publisher, publication_year, journal_business_model
ORDER BY 
  CASE WHEN cr_publisher = 'Other' THEN 1 ELSE 0 END,
  SUM(COUNT(DISTINCT doi)) OVER (PARTITION BY cr_publisher) DESC,
  publication_year DESC
```

```{r}
top_20_oa
write_csv(top_20_oa, "data/top_20_oa.csv")
```

Create publisher-level table by publication_year, oa_business_model and country

```{sql}
#| connection: bq_con
#| output.var: top_20_by_country
WITH top_20 AS (
  SELECT cr_publisher, COUNT(DISTINCT doi) AS total_articles
  FROM `subugoe-closed.apc_science.md_raw`
  GROUP BY cr_publisher
  ORDER BY total_articles DESC
  LIMIT 20
),
labeled AS (
  SELECT 
    m.doi,
    m.publication_year,
    m.oa_status,
    m.country_code,
    CASE WHEN t.cr_publisher IS NOT NULL THEN m.cr_publisher ELSE 'Other' END AS cr_publisher,
    CASE
      WHEN m.is_in_doaj = TRUE OR m.cr_publisher IN ('MDPI AG', 'Frontiers Media SA', 'Public Library of Science (PLoS)') THEN 'Full OA' ELSE 'Hybrid'
    END AS journal_business_model,
    CASE WHEN jct.doi IS NOT NULL THEN TRUE ELSE FALSE END AS enabled_ta
  FROM `subugoe-closed.apc_science.md_raw` m
  LEFT JOIN top_20 t USING (cr_publisher)
  LEFT JOIN (SELECT DISTINCT doi FROM `subugoe-closed.apc_science.jct_articles`) AS jct
    USING (doi)
)
SELECT 
  cr_publisher,
  country_code,
  publication_year,
  journal_business_model,
  COUNT(DISTINCT doi) AS n_articles, 
  COUNT(DISTINCT CASE WHEN oa_status IN ('gold', 'hybrid') THEN doi END) AS n_oa_articles,
  COUNT(DISTINCT CASE WHEN enabled_ta THEN doi END) AS n_ta_articles
FROM labeled
GROUP BY cr_publisher, country_code, publication_year, journal_business_model
ORDER BY 
  CASE WHEN cr_publisher = 'Other' THEN 1 ELSE 0 END,
  SUM(COUNT(DISTINCT doi)) OVER (PARTITION BY cr_publisher) DESC,
  publication_year DESC
```


```{r}
top_20_by_country
write_csv(top_20_by_country, "data/top_20_by_country.csv")
```

Total by country, year, and OA status, without publisher

```{sql}
#| connection: bq_con
#| output.var: total_by_country_oa
WITH labeled AS (
  SELECT
    m.doi,
    m.publication_year,
    m.oa_status,
    m.country_code,
    CASE
      WHEN m.is_in_doaj = TRUE OR m.cr_publisher IN ('MDPI AG', 'Frontiers Media SA', 'Public Library of Science (PLoS)') THEN 'Full OA' ELSE 'Hybrid'
    END AS journal_business_model,
    CASE WHEN jct.doi IS NOT NULL THEN TRUE ELSE FALSE END AS enabled_ta
  FROM `subugoe-closed.apc_science.md_raw` m
  LEFT JOIN (SELECT DISTINCT doi FROM `subugoe-closed.apc_science.jct_articles`) AS jct
    USING (doi)
)
SELECT
  country_code,
  publication_year,
  journal_business_model,
  COUNT(DISTINCT doi) AS n_articles,
  COUNT(DISTINCT CASE WHEN oa_status IN ('gold', 'hybrid') THEN doi END) AS n_oa_articles,
  COUNT(DISTINCT CASE WHEN enabled_ta THEN doi END) AS n_ta_articles
FROM labeled
GROUP BY country_code, publication_year, journal_business_model
ORDER BY publication_year DESC
```

```{r}
total_by_country_oa
write_csv(total_by_country_oa, "data/total_by_country_oa.csv")
```
