---
title: "OA by funder"
format: 
    gfm:
      echo: true
      message: false
      warning: false
execute: 
  cache: true
---

```{r}
#| echo: false
# Custom theme inspired by Our World in Data)
library(ggplot2)
my_theme <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      # Remove all grid lines except horizontal major
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      panel.grid.major.y = element_line(color = "grey85", linewidth = 0.3),
      # Clean axis styling
      axis.line.x = element_line(color = "grey40", linewidth = 0.4),
      axis.ticks.x = element_line(color = "grey40", linewidth = 0.3),
      axis.ticks.length.x = unit(0.15, "cm"),
      axis.ticks.y = element_blank(),
      # Typography
      axis.text = element_text(color = "grey30", size = rel(0.9)),
      axis.title.y = element_text(color = "grey30", size = rel(0.95), margin = margin(r = 10)),
      plot.title = element_text(face = "bold", size = rel(1.1), color = "grey10", margin = margin(b = 8)),
      plot.title.position = "plot",
      plot.caption = element_text(hjust = 1, size = rel(0.7), color = "grey50", margin = margin(t = 12)),
      plot.caption.position = "plot",
      # Legend
      legend.position = "top",
      # Margins
      plot.margin = margin(10, 10, 10, 10)
    )
}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "99%",
  fig.align = "center",
  dpi = 300,
  fig.retina = 2,
  dev = "ragg_png"
)
```

## Aim

The aim of this notebook is to demonstrate how to create a crosswalk of different Google BigQuery datasets that can be discovered through ORION-DBs. 
It would be interesting to track the development of open access over the years by country and funder. 
Of particular interest is comparing the US with European countries, since the NIH's recent proposals for supporting open access financially suggest similar funding approaches, with transformative agreements playing an increasing role in accommodating those costs and potential price caps.

## Overview of data sources used

In this notebook, we use the following sources to create a unified dataset on open access uptake across countries and funders, focusing particularly on publisher-provided open access through transformative agreements.

The data sources to be cross-walked are:

- **PubMed**: We use PubMed to retrieve NIH grant-supported publications and as an idication for biomedical literature. 
- **Open access evidence**: OpenAlex provides article- and journal-level open access information. We use the February 2026 OpenAlex snapshot provided by SUB Göttingen.
- **Publisher metadata**: Crossref is used for publisher disambiguation, as it provides more coherent publisher names without imprints compared to OpenAlex.
- **Document classification**: A document classification dataset provided by SUB Göttingen is used to restrict the sample to original research articles.
- **Articles under transformative agreements**: SUB Göttingen provides a dataset about journals, institutions, and articles that are enabled by transformative agreements.
- **Grant-supported publications**: We use the OpenAIRE graph, provided by Sesame Open Science, to obtain grant-supported publications from the NIH and the European Commission.

When conducting open access analysis, it is important to relate it to a meaningful overall sample. In this tutorial, we will focus on articles representing original research, according to a document classification approach, as they are often invoiced to institutions, and on articles which are indexed in PubMed, to restrict ourselves to life sciences literature. 
The latter can be inferred from both OpenAlex and OpenAIRE.

For country analysis, we use the first author's country affiliation from OpenAlex.

## Data gathering steps

### 1. Create publication metadata set

Initially, it is often helpful to start with a subset of publications and select only the columns needed for your analysis. 
This saves on query costs, keeps the dataset manageable, and safeguards the raw metadata in case an ORION provider updates their data.

The first step is to create a publication dataset enriched with grant funding information from the NIH and the European Commission.

NIH Funding data are retrieved from PubMed. Here, the API is used and a list of PMIDS, a databse identifier, is retrieved and then imported into BigQuery. 

European Commission funding data are obtained from the OpenAIRE graph, as provided by Sesame Open Science, and linked to publication metadata from OpenAlex and Crossref.

The query makes use of temporary tables introduced with `WITH`, also known as Common Table Expressions (CTEs). 
The first CTE extracts funded publications from OpenAIRE by unnesting the fundings array and filtering for the two funders of interest, then joining to the OpenAIRE publication and relation tables to retrieve the associated DOIs. 
This is then joined to the main publication dataset.

In the main query, we select metadata about all articles in the OpenAlex snapshot and apply a number of filters to restrict the dataset to relevant publications. The focus is on original research articles and reviews, using first author affiliations only. Note that in the OpenAlex snapshot used here, the author position field was not exported by OpenAlex, so first authors are identified using the `WITH OFFSET` position. Crossref is used as the source for publisher metadata, as it provides more coherent publisher disambiguation without imprints compared to OpenAlex. The dataset is restricted to publications from 2018 to 2025.

To query BigQuery through our Quarto notebook, we use the R `bigrquery` DBI interface. To connect:

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(bigrquery)
library(DBI)
library(countrycode)

# Connect to GBQ with billing project from SUB Göttingen, 
# You have to use your own :-)
bq_con <- dbConnect(
  bigrquery::bigquery(),
  project = "subugoe-collaborative",
  dataset = "openalex",
  billing = "subugoe-collaborative"
)
```

In the first step, we uplaod list of PMIDs to BigUqery, repreentign publicaitons associated with NIH fuinding according to PubMed.

```{r}
pmids_nih <- read_lines("nih_pmids_2018_2024.txt")
dbWriteTable(bq_con, # Our BQ connections
             "subugoe-collaborative.resources.nih_pmids_2018_2024", # Table name
             tibble::tibble(pmids_nih = pmids_nih), # PMIDs in rectangular format with column name
             overwrite = TRUE 
             )
```

In total, PubMed indexed `r length(unique(pmids_nih))` publications between 2018 and 2024 with NIH grant support.

Next, we bing it toegtehr with OpenAire, adding biomeldcial literature grant-supported by the European Commisison, and OpenAlex and Crossref to obtain schoalrly matedata about these publicaitons.

We save the table in our resource datasets for misc datasets `subugoe-collaborative.resources.oa_tutorial_md_raw`. 
You will need to define your own dataset.

```{sql}
#| connection: "bq_con"

CREATE OR REPLACE TABLE `subugoe-collaborative.resources.oa_tutorial_md_raw` AS (

WITH openaire_funders AS (
  SELECT DISTINCT
    pid.value AS pmid,
    fund.name AS funder
  FROM `sos-datasources.openaire.project_20260129` AS proj,
    UNNEST(fundings) AS fund
  INNER JOIN `sos-datasources.openaire.relation_product_project_20260129` AS pub_proj 
    ON proj.id = pub_proj.target
  INNER JOIN `sos-datasources.openaire.publication_20260129` AS pub 
    ON pub_proj.source = pub.id,
    UNNEST(pub.pids) AS pid
  WHERE fund.name IN ('European Commission')
    AND pid.scheme = 'pmid'
),

nih_funders AS (
  SELECT DISTINCT
    CAST(pmids_nih AS STRING) AS pmid
  FROM `subugoe-collaborative.resources.nih_pmids_2018_2024`
),

all_funders AS (
  SELECT pmid, 'European Commission' AS funder
  FROM openaire_funders
  UNION ALL
  SELECT pmid, 'NIH' AS funder
  FROM nih_funders
)

SELECT DISTINCT
    oalex.id,
    oalex.doi,
    oalex.ids.pmid,
    oalex.ids.pmcid,
    publication_year,
    publication_date,
    primary_location.source.issn_l AS issn_l,
    oalex.primary_location.source.host_organization_name AS oalex_publisher,
    cr.publisher AS cr_publisher,
    primary_location.source.is_in_doaj AS is_in_doaj,
    open_access.oa_status AS oa_status,
    country AS country_code,
    inst.ror,
    af.funder AS funders
  FROM `subugoe-collaborative.openalex_walden.works` AS oalex
  LEFT JOIN
    UNNEST(authorships) AS au WITH OFFSET AS pos
  LEFT JOIN
    UNNEST(au.countries) AS country
  LEFT JOIN
    UNNEST(au.institutions) AS inst  
  LEFT JOIN `subugoe-collaborative.resources.document_classification_september25` AS doctype_classifier 
    ON oalex.doi = doctype_classifier.doi
  INNER JOIN `subugoe-collaborative.cr_instant.snapshot` AS cr 
    ON LOWER(oalex.doi) = LOWER(cr.doi)
  LEFT JOIN all_funders AS af
    ON REGEXP_EXTRACT(oalex.ids.pmid, r'(\d+)$') = af.pmid
  WHERE
      pos = 0
    AND primary_location.source.type = "journal"
    AND is_paratext = FALSE
    AND oalex.type IN ('article','review')
    AND is_research IS NOT FALSE
    AND is_xpac = FALSE
    AND (
      NOT REGEXP_CONTAINS(oalex.biblio.issue, '^[a-zA-Z]')
      OR oalex.biblio.issue IS NULL)
    AND (
      NOT REGEXP_CONTAINS(oalex.title, '[0-9]{3} pp.'))
    AND publication_year BETWEEN 2018 AND 2025
)
```

### 2. Estimate articles from transformative agreements

Next, we will estimate open access articles covered by a transformative agreement using a method described in @Jahn_2025

The following query uses several datasets from the Journal Checker tool, provided by SUB Göttingen in the OpenBib collection.

```{sql}
#| connection: "bq_con"
CREATE OR REPLACE TABLE `subugoe-collaborative.resources.oa_tutorial_jct_articles` AS (
WITH
  -- Enrich ROR variants from associated institutions
  obtain_associated_ror_ids AS (
    SELECT
      esac_id,
      jct_inst.ror_id AS ror_jct,
      inst.ror AS ror_associated
    FROM
      `subugoe-collaborative.openbib.jct_institutions` AS jct_inst
    LEFT JOIN
      `subugoe-collaborative.openalex.institutions` AS oalex_inst
    ON
      jct_inst.ror_id = oalex_inst.ror
    LEFT JOIN
      UNNEST(oalex_inst.associated_institutions) AS inst
  ),

  create_matching_table AS (
    SELECT esac_id, 'ror_jct' AS ror_type, ror_jct AS ror
    FROM obtain_associated_ror_ids
    UNION ALL
    SELECT esac_id, 'ror_associated' AS ror_type, ror_associated AS ror
    FROM obtain_associated_ror_ids
  ),

  enriched_ror_variants AS (
    SELECT DISTINCT
      create_matching_table.*,
      DATE(jct_inst.start_date) AS start_date,
      DATE(jct_inst.end_date) AS end_date
    FROM create_matching_table
    INNER JOIN `subugoe-collaborative.openbib.jct_esac` AS jct_inst
      ON create_matching_table.esac_id = jct_inst.id
  ),

  journal_agreements AS (
    SELECT
      j.issn_l,
      e.id AS esac_id
    FROM `subugoe-collaborative.openbib.jct_journals` AS j
    INNER JOIN `subugoe-collaborative.openbib.jct_esac` AS e
      ON j.esac_id = e.id
  )

SELECT DISTINCT
  md.id,
  md.doi,
  md.issn_l AS matching_issn_l,
  md.ror AS matching_ror,
  erv.ror_type,
  erv.esac_id,
  erv.start_date,
  erv.end_date,
  md.publication_date
FROM `subugoe-collaborative.resources.oa_tutorial_md_raw` AS md
INNER JOIN journal_agreements AS ja
  ON md.issn_l = ja.issn_l
INNER JOIN enriched_ror_variants AS erv
  ON md.ror = erv.ror
  AND ja.esac_id = erv.esac_id
WHERE
  md.ror IS NOT NULL
  AND md.issn_l IS NOT NULL
  AND (PARSE_DATE('%Y-%m-%d', md.publication_date) >= erv.start_date OR erv.start_date IS NULL)
  AND (PARSE_DATE('%Y-%m-%d', md.publication_date) <= erv.end_date OR erv.end_date IS NULL)
  -- Only TA OA articles
  AND oa_status IN ('gold', 'hybrid')
ORDER BY erv.esac_id, md.publication_date
)
```

With these two datasets compiled, we are ready to proceed with the open access analysis.

### Create analytical dataset: Funder view

```{sql}
#| connection: "bq_con"
#| output.var: funding_oa
WITH
  oa AS (
    SELECT
      publication_year,
      funders,
      CASE
        WHEN (is_in_doaj = TRUE AND oa_status = 'gold') THEN 'gold_is_in_doaj'
        WHEN (is_in_doaj = TRUE AND oa_status = 'diamond')
          THEN 'diamond_is_in_doaj'
        WHEN oa_status = 'hybrid' THEN 'hybrid_oa'
        WHEN oa_status != 'closed' THEN 'other_oa'
        END AS oa_type,
      CASE
        WHEN jct.doi IS NOT NULL THEN TRUE
        ELSE FALSE
        END AS enabled_ta,
      COUNT(DISTINCT md.pmid) AS n_oa_articles
    FROM `subugoe-collaborative.resources.oa_tutorial_md_raw` AS md
    LEFT JOIN
      (
        SELECT DISTINCT doi
        FROM `subugoe-closed.apc_science.jct_articles`
      ) AS jct
      USING (doi)
    WHERE funders IS NOT NULL
    GROUP BY
      publication_year,
      funders,
      oa_type,
      enabled_ta
  ),
  total AS (
    SELECT
      publication_year,
      funders,
      COUNT(DISTINCT pmid) AS articles
    FROM `subugoe-collaborative.resources.oa_tutorial_md_raw`
    GROUP BY
      publication_year,
      funders
  )
SELECT
  total.publication_year,
  total.funders,
  total.articles,
  oa.n_oa_articles,
  oa.oa_type,
  oa.enabled_ta
FROM total
LEFT JOIN oa
  ON
    total.publication_year = oa.publication_year
    AND total.funders = oa.funders
WHERE total.funders IS NOT NULL AND total.publication_year < 2025
ORDER BY total.publication_year DESC
```

Her eis the table, which Is tored in an R object `funding_oa`for further local data analysis.
```{r}
funding_oa
```

```{r}
oa_by_year <- funding_oa |>
  filter(!is.na(oa_type)) |>
  summarise(n_oa = sum(n_oa_articles),
            .by = c(publication_year, funders, articles)
            ) |>
  mutate(oa_prop = n_oa / articles)
```

```{r}
#| dpi: 300
#| fig-retina: 2
#| label: oa_by_year_rel
oa_by_year |>
  ggplot(aes(publication_year, oa_prop, group = funders, color = funders)) +
  geom_line() + 
  geom_point() +
  scale_y_continuous("Open Access (in%)", labels = scales::percent_format()) +
  scale_color_manual(NULL,
    values = c(
      "NIH" = "#323a45",
      "European Commission" = "#F8AE21"
    )) +
  labs(x = NULL, 
       title = "Open Access to biomedical research literature",
       subtitle = "Journal articles and reviews in PubMed",
       caption = "Data sources: PubMed, OpenAIRE, OpenAlex.") +
  my_theme()
```

```{r}
#| dpi: 300
#| fig-retina: 2
#| label: oa_by_year_abs
oa_by_year |>
  ggplot(aes(publication_year, n_oa, group = funders, color = funders)) +
  geom_line() + 
  geom_point() +
  scale_y_continuous("Open Access", labels = scales::label_number(big.mark = ","), limits = c(0, NA), expand = c(0, 0)) +
  scale_color_manual(NULL,
    values = c(
      "NIH" = "#323a45",
      "European Commission" = "#F8AE21"
    )) +
  labs(x = NULL, 
       title = "Open Access to biomedical research literature",
       subtitle = "Journal articles and reviews in PubMed",
       caption = "Data sources: PubMed, OpenAIRE, OpenAlex.") +
  my_theme()
```

Open access by open access business model. we distinguish between

APC: gold and hybrid not via TA
TA: gold and hybrid via TA
Diamond: diamond
Other 

```{r}
oa_by_year_and_type <- funding_oa |>
  filter(!is.na(oa_type)) |>
  mutate(oa_cat = case_when(
    oa_type %in% c("gold_is_in_doaj", "hybrid_oa") & enabled_ta == FALSE ~ "APC",
    oa_type %in% c("gold_is_in_doaj", "hybrid_oa") & enabled_ta == TRUE ~ "Transformative Agreement",
    oa_type == "diamond_is_in_doaj" ~ "Diamond OA",
    .default = "Other")
) |> summarise(n_oa = sum(n_oa_articles),
            .by = c(publication_year, funders, articles, oa_cat)
            ) |>
  mutate(oa_prop = n_oa / articles,
             oa_cat = factor(oa_cat, levels = c("APC", "Transformative Agreement", "Diamond OA", "Other")))
```


```{r}
oa_by_year_and_type
```

Relative

```{r}
#| fig-width: 8
#| dpi: 300
#| fig-retina: 2
#| label: oa_by_year_type_rel
oa_by_year_and_type |>
  ggplot(aes(publication_year, oa_prop, fill = fct_rev(oa_cat))) +
  geom_area(color = "black", linewidth = 0.2) +
  facet_wrap(~funders) +
  scale_y_continuous("Open Access (in%)", labels = scales::percent_format(), limits = c(0, 1),
                   expand = c(0, 0)) +
  scale_fill_manual("OA model",
    values = c(
      "APC" = "#0E6BA0",
      "Transformative Agreement" = "#30C0D2",
      "Diamond OA" = "#E2A3A4",
      "Other" = "#AEACA0"
    )) +
  labs(x = NULL, 
       title = "Open Access to grant-supported biomedical research literature",
       subtitle = "Journal articles and reviews in PubMed",
       caption = "Data sources: PubMed, OpenAIRE, OpenAlex, cOAlition S Journal Checker Tool.") +
  my_theme() +
  guides(fill = guide_legend(reverse = TRUE))

```

Absolute

```{r}
#| fig-width: 8
#| dpi: 300
#| fig-retina: 2
#| label: oa_by_year_type_abs

oa_by_year_and_type |>
  ggplot(aes(publication_year, n_oa, fill = fct_rev(oa_cat))) +
  geom_area(color = "black", linewidth = 0.2) +
  facet_wrap(~funders) +
  scale_y_continuous("Open Access", labels = scales::label_number(big.mark = ","), 
                     limits = c(0, NA), expand = c(0, 0)) +
  scale_fill_manual("OA model",
    values = c(
      "APC" = "#0E6BA0",
      "Transformative Agreement" = "#30C0D2",
      "Diamond OA" = "#E2A3A4",
      "Other" = "#AEACA0"
    )) +
  labs(x = NULL, 
       title = "Open Access to grant-supported biomedical research literature",
       subtitle = "Journal articles and reviews in PubMed",
       caption = "Data sources: PubMed, OpenAIRE, OpenAlex, cOAlition S Journal Checker Tool.") +
  my_theme() +
  guides(fill = guide_legend(reverse = TRUE))

```

Key figures 2024

```{r, echo=FALSE}
library(tidyverse)
options(scipen = 999, digits = 1)
knitr::opts_chunk$set(format.args = list(big.mark = ",", digits = 2))

```

NIH

- `r oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(oa_prop) |> sum() *100`% of grant-supported biomedical research articles were open access (`r oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(n_oa) |> sum()` out of `r oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(articles) |> unique()` original research articles and reviews indexed in PubMed and OpenAlex).

- `r oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024, oa_cat == "APC") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(n_oa) |> sum()* 100`% of open access articles were enabled through APCs, and `r oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024, oa_cat == "Transformative Agreement") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(n_oa) |> sum() * 100`% through transformative agreements. `r  oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024, oa_cat == "Other") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "NIH", publication_year == 2024) |> pull(n_oa) |> sum() * 100` was exclusively open access through other means, particularly through PubMed Central.

European Commission

- `r oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(oa_prop) |> sum() *100`% of grant-supported biomedical research articles were open access (`r oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(n_oa) |> sum()` out of `r oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(articles) |> unique()` original research articles and reviews indexed in PubMed and OpenAlex).

- `r oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024, oa_cat == "APC") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(n_oa) |> sum()* 100`% of open access articles were enabled through APCs, and `r oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024, oa_cat == "Transformative Agreement") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(n_oa) |> sum() * 100`% through transformative agreements. `r  oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024, oa_cat == "Other") |> pull(n_oa) / oa_by_year_and_type |> filter(funders == "European Commission", publication_year == 2024) |> pull(n_oa) |> sum() * 100`% was exclusively open access through other means.
